Competing Bandits: The Perils of Exploration under Competition

### 2018-09-02 paper/content/inverted_u.tex
fixups

Since both firms enter at the same time the game is symmetric and the competition game is zero-sum so we only need to compute the expected payoffs of three pairs of strategies. We do not report them in the figures, but when both firms play the same algorithm the expected market share is 50/50. Since our results are from simulations, we interpret algorithm $A$ being better than algorithm $B$ if the resulting market share of $A$ is statistically greater than $0.5$. Otherwise, the firm is indifferent between $A$ and $B$ with ties broken towards algorithms that are easier to deploy.

@@

\begin{puzzle}\label{puz:dg_ht}
Even for large warm start, $\DG$ is better than $\DEG$ under Heavy Tail
\end{puzzle}

>>>>

Since both firms enter at the same time the game is symmetric and the competition game is zero-sum so we only need to compute the expected payoffs of three pairs of strategies. We do not report them in the figures, but when both firms play the same algorithm the expected market share is 50/50. Since our results are from simulations, we interpret algorithm $A$ being better than algorithm $B$ if the resulting average market share of $A$ is statistically greater than $0.5$. Otherwise, the firm is indifferent between $A$ and $B$ with ties broken towards algorithms that are easier to deploy.

@@

\begin{puzzle}\label{puz:dg_ht}
Even for large warm start, $\DG$ is better than $\DEG$ under Heavy Tail \footnote{We demonstrate this holds even for $K = 3$ in the supplementary material}
\end{puzzle}

### 2018-09-02 paper/content/revisited.tex
fixups

Figure \ref{relative_rep_plots} shows plots of the relative reputation proportion for $\TS$ vs $\DG$ on the Uniform and Needle In Haystack instances. For the Uniform instance we see that, in the early rounds, the relative reputation proportion for $\TS$  vs. $\DG$ is less than 0.5 meaning that $\DG$ has a higher reputation in a majority of the simulations  but that, eventually, the proportion $\TS$ does better than $\DG$ in a majority of the simulations. The intuition behind this is that, especially since the firms start with no substantive initial information, $\TS$ does purposeful exploration in the early rounds in order to acquire information. However, eventually the information acquired from purposeful exploration in the early rounds allows $\TS$ to make better decisions and achieve a higher reputation, especially when the instance is ``hard enough" so that $\DG$ cannot trivially find the best arm. The early exploration leads to what we define as a \textit{relative reputation cost} and the eventual gain in reputation leads to what we define as a \textit{relative reputation benefit}.

\begin{definition}
\textit{Relative reputation cost and benefit} - the relative reputation loss an algorithm incurs from purposeful exploration compared to the greedy alternative. Here we treat it as an empirical definition based on the relative reputation proportion where the ``cost" regime is when the proportion $< 0.5$ and the ``benefit" regime is when the proportion is $> 0.5$ for the better algorithm. We call the priors where there is an early costly period followed by a later benefit period relative reputation costly instances.
\end{definition}

Is exploration always costly? Figure \ref{relative_rep_plots} also shows that for the Needle In Haystack instances, $\TS$ always does relatively better than $\DG$. There are two contributing factors to this. First, $\TS$ identifies the best arm faster in the Needle In Haystack prior than the Uniform prior so that there is a shorter time horizon where $\TS$ needs to engage in purposeful exploration. Second, in the Needle In Haystack prior there are no ``bad" arms as there may be in the Uniform prior since by construction all the arms except one in Needle In Haystack are the same. Thus, when $\TS$ pulls a sub-optimal arm relative to its current information, the expected reward is the same as the greedy option that has not identified the best arm. However, with the Uniform prior, it is possible that the sub-optimal arm that is pulled has substantially lower expected reward relative to the greedy option. Thus, only the Uniform and Heavy Tail instances are relative reputation costly priors.

@@

Thus, the reason we see that $\DG > \TS$ under Heavy Tail and Uniform but that $\TS$ is always preferred under Needle In Haystack is due to the fact that the purposeful exploration engaged by $\TS$ in the early rounds leads it to suffer a relative reputational cost in the former case but not in the latter. Looking at the relative reputation plots in Figure \ref{relative_rep_plots}, we can interpret fixing a warm start $T_0$ as fixing the starting point on the relative reputation plots. The proportion of first rounds in the competition game that will go to a firm playing alg $A$ over a firm playing alg $B$ will correspond to the relative reputation proportion at time $T_0$. As a result, if we begin with a warm start that falls in the region where the proportion $< 0.5$, we expect that $\DG > \TS$.

This also allows us to give some intuition as to why allowing one firm to be a temporary monopoly incentivizes it to play $\TS$ even if, for the same warm start, the firm would be incentivized to play $\DG$ under permanent duopoly.  By allowing it to be the only firm in the market for sufficiently long we allow it to recoup the ``reputation costs" by having it engage in most of its exploration while it does not have the pressure to incentivize agents to select them over their competition.

>>>>

Figure \ref{relative_rep_plots} shows plots of the relative reputation proportion for $\TS$ vs $\DG$ on the Uniform and Needle In Haystack instances. For the Uniform instance we see that, in the early rounds, the relative reputation proportion for $\TS$  vs. $\DG$ is less than 0.5 meaning that $\DG$ has a higher reputation in a majority of the simulations  but that, eventually, $\TS$ does better than $\DG$ in a majority of the simulations. The intuition behind this is that, especially since the firms start with no substantive initial information, $\TS$ does purposeful exploration in the early rounds in order to acquire information which leads to a relatively lower reputation compared to $\DG$. However, eventually the information acquired from purposeful exploration in the early rounds allows $\TS$ to make better decisions and achieve a higher reputation, especially when the instance is ``hard enough" so that $\DG$ cannot trivially find the best arm. The early exploration leads to what we define as a \textit{relative reputation cost} and the eventual gain in reputation leads to what we define as a \textit{relative reputation benefit}.

\begin{definition}
\textit{Relative reputation cost and benefit} - the relative reputation loss an algorithm incurs from purposeful exploration compared to the greedy alternative. Here we treat it as an empirical definition based on the relative reputation proportion where the ``cost" regime is when the proportion $< 0.5$ and the ``benefit" regime is when the proportion is $> 0.5$ for the better algorithm. We call the instances where there is an early costly period followed by a later benefit period relative reputation costly instances.
\end{definition}

Is exploration always costly? Figure \ref{relative_rep_plots} also shows that for the Needle In Haystack instances, $\TS$ always does relatively better than $\DG$. There are two contributing factors to this. First, $\TS$ identifies the best arm faster in the Needle In Haystack instances than the Uniform instances so that there is a shorter time horizon where $\TS$ needs to engage in purposeful exploration. Second, in the Needle In Haystack instances there are no ``bad" arms as there may be in the Uniform instances since by construction all the arms except one in Needle In Haystack are the same. Thus, when $\TS$ pulls a sub-optimal arm relative to its current information, the expected reward is the same as the greedy option that has not identified the best arm. However, with the Uniform instances, it is possible that the sub-optimal arm that is pulled has substantially lower expected reward relative to the greedy option. Thus, only the Uniform and Heavy Tail instances are relative reputation costly instances.

@@

Thus, the reason we see that $\DG$ is better than $\TS$ under Heavy Tail and Uniform but that $\TS$ is always preferred under Needle In Haystack is due to the fact that the purposeful exploration engaged by $\TS$ in the early rounds leads it to suffer a relative reputational cost in the former case but not in the latter. Looking at the relative reputation plots in Figure \ref{relative_rep_plots}, we can interpret fixing a warm start $T_0$ as fixing the starting point on the relative reputation plots. The proportion of first rounds in the competition game that will go to a firm playing alg $A$ over a firm playing alg $B$ will correspond to the relative reputation proportion at time $T_0$. As a result, if we begin with a warm start that falls in the region where the proportion $< 0.5$, we expect that $\DG$ does better than $\TS$.

This also allows us to give some intuition as to why allowing one firm to be a temporary monopoly incentivizes it to play $\TS$ even if, for the same warm start, the firm would be incentivized to play $\DG$ under permanent duopoly.  By allowing it to be the only firm in the market for sufficiently long we allow it to recoup the reputation costs by having it engage in most of its exploration while it does not have the pressure to incentivize agents to select them over their competition.
